Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material

### 2019-05-23 paper/introduction.tex
Typos

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g. words that contain "z" are rare) and so forth \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.

@@

We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}). After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g. words that contain `z' are rare) and so forth \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.

@@

We introduce the `back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}). After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/meg_exps.tex
Typos

For each subject and each time sample relative to word onset, we
build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words
by $q=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

\subsubsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using Spacy \citep{spacy2} and
WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency). This
procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
$d=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

@@

Forward regression was performed with an l2-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
%
Backward regression was performed with an l2-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
%
Finally, we used the back-to-back regression was performed using two l2-regularized
least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.

>>>>

For each subject and each time sample relative to word onset, we
build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2700 words
by $d_y=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

\subsubsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using Spacy \citep{spacy2} and
WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency). This
procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
$d_x=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

@@

Forward regression was performed with an $\ell_2$-regularized least-squares regression $H \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| H_{i, :} \|^2$.
%
Backward regression was performed with an $\ell_2$-regularized least-squares regression $G \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YG$, using a validation sample.
%
Finally, we used the back-to-back regression was performed using two $\ell_2$-regularized
least squares regressions $G$ and $H$ as described in section \ref{sec:algorithm}, using a bagging of 10 alphas logarithmically distributed between -5 and 5.

### 2019-05-23 paper/paper.tex
Typos

    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times q}$}};
    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d}$}};
    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d \times d}$}};
    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d}$}};
    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d \times q}$}};

@@

Back-to-Back Regression (B2B) consists of two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{E} \leftarrow \hat{E} + \text{Diag}(H)$\;
    }
    $\hat{E} \leftarrow \hat{E} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
    \Return{$\hat{E}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
%
On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
%
In this section, we provide a recipe to binarize $\hat{E}$ and estimate the collection of causal factors.

In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

@@

We perform two experiments to evaluate the performance of B2BR: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
%
We use scikit-learn to implement all of our simulations \citep{sklearn}.

>>>>

    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};

@@

Back-to-Back Regression (B2B) consists of two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(\hat{H})$, is the desired estimate of the causal influence matrix $E$.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{E} \leftarrow \hat{E} + \text{Diag}(\hat{H})$\;
    }
    $\hat{E} \leftarrow \hat{E} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
    \Return{$\hat{E}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
%
On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
%
In this section, we provide one of the many possible recipes to binarize $\hat{E}$ and estimate the collection of causal factors.

In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.

@@

We perform two experiments to evaluate the performance of B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
%
We use scikit-learn to implement all of our simulations \citep{sklearn}.

### 2019-05-23 paper/synthetic.tex
Typos

The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $snr$ is a scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize}
%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%\end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
and the factor $\text{snr} \in (0, \infty)$.

@@

We evaluate seven metrics:
%
test error in-distribution, test error out-distribution, test error-in distribution after weighting with $\hat{E}$, test error out-distribution after weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$, Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve (AUC) between $\hat{E}$ and $E$.
%
The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated mask $\hat{E}$.
%
The ``out-distribution'' metrics are similar, measuring the test error at samples drawn using two different random covariance matrices for both $X$ and $N$.
%
The false positives, false negatives, and AUC statistics measure the quality of identification of causal factors between $\hat{E}$ and the true, binary $E$.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, $20$ random seeds.}
    \label{table:synthetic}
\end{figure}

Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
%
B2B is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.

>>>>

The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize}
%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%\end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
and the factor $\text{snr} \in (0, \infty)$.

@@

We evaluate seven metrics:
%
test error in-distribution, test error out-distribution, test error-in distribution after weighting with $\hat{E}$, test error out-distribution after weighting with $\hat{E}$, Sonquist-Morgan false positives on $\hat{E}$, Sonquist-Morgan false negatives on $\hat{E}$, and the area under the ROC curve (AUC) between $\hat{E}$ and $E$.
%
The ``in-distribution'' metrics measure the test error of each predictor before and after weighting their features by the estimated (continous!) mask $\hat{E}$.
%
The ``out-distribution'' metrics are similar, measuring the test error at samples drawn using two different random covariance matrices for both $X$ and $N$.
%
The false positives, false negatives, and AUC statistics measure the quality of identification of causal factors between $\hat{E}$ and the true, binary $E$.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.}
    \label{table:synthetic}
\end{figure}

Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
%
B2B is the method obtaining best test errors, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.
